Figures to insert in text

### 2018-03-10 Chapters/Chapter02.tex
SDM Splitting into multiple files.

\chapter{Sparse Distributed Memory}

\input{./Chapters/sdm2.tex}

\chapter{Framework Architecture}

The framework implements the basic operations in a Sparse Distributed Memory which may be used to create more complex operations. It is developed in C language and the OpenCL parallel framework --- which may be loaded in many platforms and programming languages --- with a wrapper in Python. The Python module makes it easy to create and execute simulations in a Sparse Distributed Memory and works properly in Jupyter Notebook \citep{kluyver2016jupyter}. It works in both Python 2 and Python 3.

We split the SDM memory in two parts: the hard-location addresses and the hard-location counters. Thus, the addresses (bitstrings) of the hard-locations are stored in one array, while their counters in another. This makes possible to create multiple SDMs using the same address space, which would save computational effort to scan a bitstring in all the SDMs --- since they share the same address space, the activated hard-locations will be the same in all of them. As the slowest part of reading and writing operations is scanning the address space, the performance benefits are significant.

Each part may be stored either in the RAM memory or in a file. The RAM memory is interesting for quick experiments, automated tests, and others scenarios in which the SDM may be lost, while the file is interesting for a long-term SDM, like creating an SDM file with 10,000 random writes, which will be copied over and over to run multiple experiments. The file may also be sent to another researcher or may be published within the paper to let others run their own checks and verify the results. In summary, the framework fits many different uses and necessities.

Let a SDM memory with $N$ dimensions and $H$ hard-locations. Then, in a 64-bit computer, the array of hard-location addresses will use $H \cdot 8 \cdot \lceil N/64 \rceil$ bytes of memory, and there will be $H \cdot N$ hard-location counters. For example, in a SDM memory with 1,000 dimensions and 1,000,000 hard-locations, using 32-bit integers for the counters, the array of addresses will use 122MB of memory and the counters will use 3.8 GB of memory.

Basic operations were grouped in four sets: (i) for bitstrings, (ii) for addresses, (iii) for counters, and (iv) for memories (SDMs). Operations include creating new bitstrings, flipping bits, generating a bitstring with a specific distance from a given bitstring, scanning the address space using different algorithms, writing a bitstring to a counter, writing in an SDM, reading from an SDM, and iteratively reading from an SDM until convergence.

\section{Bitstring}

Bitstrings are the main structure of SDM. The addresses are represented in bitstrings, as well as the data. A bitstring is stored as an array of integers. Each integer may be 16-bit, 32-bit, or 64-bit long, depending on the configuration. By default, each integer is 64-bit long.

For instance, a 1,000-bit bitstring will have $\lceil 1000/64 \rceil = 16$ integers. These integers will have a total of $16 \cdot 64 = 1,024$ bits. The remaining 24 bits are always zero, so they do not affect the result of any operation. The memory usage efficiency is $1 - 24/1024 = 97.65\%$. Bitstrings store neither how many bits they have nor the array length. These pieces of information are only stored in the address space.

\subsection{The distance between two bitstrings}

The distance between two bitstrings is calculated by the hamming distance, which is the number of different bits between them. It is calculated counting the number of ones in the exclusive or (xor) between the bitstrings, i.e., $d(x, y) = \text{number of ones in } x \oplus y$.

There are several algorithms to calculate the number of ones \citep{warren2013hacker}, but the performance depends on the processor. So, we have implemented three different algorithms and one may be selected through compiling flags. The default algorithm is to use a built-in \_\_popcnt() instruction from the compiler.

There is also the naive algorithm, which really counts the number of ones checking bit by bit. It is available only to testing purposes and should never be used.

The other algorithm available is the lookup. It pre-calculates a table with the number of ones of all possible 16-bit integers. This table is accessed a few times to calculate the number of ones of a 64-bit integer, i.e., to calculate the distance between two bitstrings, it sums the distance of each 16-bit part of the bitstrings, i.e., $d(x[0:63], y[0:63]) = d(x[0:15], y[0:15]) + d(x[16:31], y[16:31]) + d(x[32:47], y[32:47]) + d(x[48:63], y[48:63])$ where $x[i:i+15]$ and $y[i, i+15]$ are the 16-bit integers formed by the bits between $i$ and $i+15$ of $x$ and $y$, respectivelly. Each 16-bit distance is calculated through a single table access. As each distance is calculated in O(1), this algorithm runs in O($\lceil bits/16 \rceil$). This table uses 65MB of RAM. One may change the table from 16-bit integers to 32-bit integers, which would halve the number of accesses at the expense of 4GB of RAM (instead of 65MB).

\section{Address space}

An address space is a fixed collection of bitstrings, and each bitstring represents a hard-location address. They store the number of bitstrings, as well as the number of bits, number of integers per bitstring, and the number of remaining bits.

Bitstrings are stored in a contiguous array of 64-bit integers, as shown in Figure \ref{tab:hl-addresses-detail}. Hence, basic pointer arithmetic provides us with performance improvements in their access, as processors realize fetches of contiguous chunks of memory  \citep{pai2004linux}.

\begin{figure}
\centering
\begin{tikzpicture}[
mycell/.style={draw, minimum size=7mm},
matrixA/.style={matrix of nodes,
    nodes={mycell, anchor=center},
    column sep=-\pgflinewidth,
    row sep=-\pgflinewidth,
    },
matrixB/.style={matrix of nodes,
    nodes={mycell, anchor=center},
    column sep=-\pgflinewidth,
    row sep=-\pgflinewidth,
}]

\matrix[matrixA] (A) { addr$_1$ & addr$_2$ & addr$_3$ & $\cdots$ & addr$_H$ \\ };

\matrix[matrixB, below=of A] (B) {
addr$_{k, 1}$ & addr$_{k, 2}$ & addr$_{k, 3}$ & $\cdots$ & addr$_{k, 8 \cdot \lceil N/64 \rceil}$ \\
};

\draw[dashed] (A-1-1.south west)--(B-1-1.north west);
\draw[dashed] (A-1-1.south east)--(B-1-5.north east);
\draw [
	thick,
    decoration={
        brace,
        mirror,
		amplitude=0.2cm,
        raise=0.2cm
    },
    decorate
] (B-1-1.south west) -- (B-1-5.south east)
node [pos=0.5,anchor=north,yshift=-0.5cm] {N bits};

\end{tikzpicture}

\caption{Address space's bitstrings are stored in a contiguous array. In a 64-bit computer, each bitstring is stored in a sub-array of 64-bit integers, with length $8 \cdot \lceil N/64 \rceil$.\label{tab:hl-addresses-detail}}
\end{figure}

The scan for activated hard-locations is performed in an address space. It returns the indexes of the bitstrings which were inside the circle (and their distances). Then, each operation uses these pieces of information in a different way.

\subsection{Scanning for activated hard-locations}

Scanning for the activated hard-locations is a problem similar to well-known problems in computational geometry called ``range reporting in higher dimensions''. In this case, none of the known algorithms is able to solve our problem faster than $O(H)$. The algorithm which seems to best fit in our problem consumes $O(H)$ space and runs in $O(\log^n(H))$ \citep{chazelle1988functional}, which is really slower than $O(H)$ when, for instance, $H=1,000,000$ and $n=1,000$. For a review of the range reporting algorithms, see \citet{chan2011orthogonal}.

In 2014, there was published a solution to fast search in hamming space which seems applicable to our problem \citet{norouzi2014fast}. It provides a fast search when $r/n < 0.11$ or $r/n < 0.06$, where $r$ is the radius and $n$ is the number of bits. But, in our case, for a 1,000 bits SDM, $r/n = 0.451$, which changes the runtime to $O(H^{0.993})$. This is really close to $O(H)$, but with a larger constant. Unfortunately, $O(H)$ is still faster.

It is intriguing that none of those algorithms is able to solve our scanning problem. The idea behind those computational geometry algorithms is roughly to split the search space in half each step, which would take $O(\log(H))$ to go through the whole space. But this approach does not work because of the high number of dimensions (i.e., 1,000) and because the hard-locations' addresses are randomly sampled from the $\{0, 1\}^n$ space. Although each addresses' bit itself splits the hardlocations in half, it does not split the search space in half since both halves still must be convered by the algorithm. For instance, let's say we have $n=1,000$ dimensions with $H=1,000,000$ hard-locations, and we are scanning within a circle with radius $r=451$, then after checking the first bit we have two cases: (i) for the half with the same first bit, we must keep scanning with radius 451; and (ii) for the half with a different first bit, we must keep scanning with radius 450. Hence, the search space has not been split in half because both halves have been covered (and one of them should have been skipped).

Finally, as our best approach is to scan through all hard-locations, we may distribute the scan into many tasks which will be executed independently. The tasks may be executed in different processes, threads, or even computers. They may also run in the CPU or in the GPU. In this case, we may take into account both the time required to distribute the tasks and the time to receive their results.

The framework implements three main scanner algorithms: linear scanner, thread scanner, and OpenCL scanner. The linear scanner runs in a single core, is the slowest one, and was developed only for testing purposes; the thread scanner runs at the CPU in multiple threads sharing memory (and our recommendation is to use the number of threads equals to twice the number of CPU cores); and the OpenCL scanner runs in multiple GPU cores and support multiple devices. The speed of a scan depends on the CPU and GPU devices, thus the best approach to choose which scanner is best for one's setup is to run a benchmark.

The OpenCL must be initialized, which just copies the address space's bitstrings to the GPU's memory. Then, many scans may be executed with no necessity to upload the bitstrings again. The OpenCL scanner supports running into multiple devices.

\subsection{OpenCL kernel}

The OpenCL kernel explores the GPU architecture to improve performance. Each work group calculates the distance of several bitstrings. During the distance calculation, each worker calculates the exclusive OR (XOR) between two 64-bit integers and use the built-in popcount function to count the number of ones. Then, they update an array of intermediate distances with their partial distances. This array is stored in the local memory and is shared with all workers of the same group. This whole step happens simultaneously in the GPU. Then, we used a reduction algorithm to sum the intermediate distances array in order to calculate the correct distance. This reduction algorithm is also distributed between the workers and runs in $O(\log_2(\text{bs\_step}))$. Finally, one of the workers checks whether the distance is less than or equal to the radius to include the bitstring index into the resulting array.

The number of workers in each group is the closest power of two above bs\_len (which is the number of 64-integers that forms a bitstring).

\section{Counters}

Each hard-location has one integer of data per bit. For instance, each hard-location of a 1,000 bits SDM has 1,000 bits. Those integers are stored in a counter.

A counter is an array of integers which stores the data of all hard-locations. So, the counter's array has $n \cdot H$ integers.

When two counters are added in a third counter, there may occur an overflow. It is not supposed to be a problem because, by default, each counter is a signed 32-bit integer that can store any number between -2,147,483,648 and 2,147,483,647, which means they will not overflow with less writes than $2^{31}-1$ divided by the average number of activated hard-locations. For instance, when $n=1,000$, $H=1,000,000$, and $r=451$, the average number of activated hard-locations is 1,000 and it would require at least one million writes before being possible to a counter to overflow.  Note also that it would be more likely to saturate the memory before any overflow.

Anyway, counters may have overflow protection depending on compiling options. By default, there is no overflow check for performance reasons (and because it does not seem necessary).

\section{Read and write operations}

The reading and writing operations are executed in two steps: first, the address space is swept looking for the activated addresses; then, the operation is performed in the counters. Reading operation assemblies the bitstring according to the counters of the activated addresses, while the writing operation changes the counters.

The iterated reading keeps reading until it gets exactly the same bitstring (or the number of maximum interations has been reached), i.e., it performs $\eta_{i+1} = \text{read}(\eta_i)$ and stops when $\eta_{k+1} = \eta_{k}$. If the initial bitstring is inside the critical distance of $\eta$, it will converge to $\eta$, but, if it is not, it will diverge and reach the maximum number of iterations.

The framework has both Kanerva's read and Murilo's generalized read. The generalization brings a parameter $z$, which is the exponent. In this case, the results are floating point instead of integer, which considerably reduces performance. When $z=1$, it is exactly as the Kanerva's read. When $z=0$, it is the Chada's read. We also explored how SDM would behave for different values of $z$.

There is another special read operation: the weighted reading. In the weighted reading, the value of the counters are multiplied by a weight which depends only on the distance between the reading address and the hard-location address. The weight is retrieved from a lookup table of integers indexed by the distance. The rest of the read operation is exactly the same.

There is also a weighted writing operation. In this case, the weight is applied when the counters are updated, i.e., if the weight is 2, the counters are increased twice when bits are 1, and decreased twice when bits are 0. Just as in the weighted reading, the weights depend only on the distance between the writing address and the hard-location address. The weights are retrieved from a lookup table of integers indexed by the distance.

\chapter{Results (i): Framework Validation}

The framework has been validated comparing its results with the expected results from \citet{Kanerva1988}. Thus, we run simulations which were then compared to the theoretical analysis conducted some decades ago.

\section{Distance between random bitstrings}

As showed by \citet{Kanerva1988}, the distance between two bitstrings follows a binomial distribution with mean $\mu = n/2$ and standard deviation $\sigma = \sqrt{n}/2$. For large values of $n$, it may be approximated by a normal distribution with the same mean and standard deviation.

In order to validate our random bitstring generation algorithm, we have calculated 10,000 distances between two random bitstrings with $n=1,000$ bits. In total, 20,000 random bitstrings have been generated during the simulation. The code is available in the ``Distance between bitstrings'' notebook \citep{sdmframework}.

In figure \ref{fig:validation-distance}, we can notice that the theoretical model and the simulation matches. Hence, it seems the random bitstring generation algorithm works properly.

This also validates the algorithm used to calculate the distance between two bitstrings. In this simulation, we have used the built-in \_\_popcnt() function.

\begin{figure}[!htb]
  \centering
  \subfloat[Full histogram ]{\includegraphics[width=0.5\textwidth]{./images02/new-images/bs-hist-full.png}}
  \subfloat[{Zoom in the interval $[400, 600]$} ]{\includegraphics[width=0.5\textwidth]{./images02/new-images/bs-hist-zoom.png}}

  \caption{Histogram of 10,000 distances between two random bitstrings with 1,000 bits. The curve in red is the theoretical normal distribution with $\mu = 500$ and $\sigma = \sqrt{500}/2$.}
  \label{fig:validation-distance}
\end{figure}

\section{Number of activated hard-locations}

In his seminal work, Kanerva proposed to use a sample of 1,000,000 hard-locations in a 1,000 bits SDM. He also proposed to activate only 1,000 of them, on average. He calculated that an access radius of $r=451$ would activate, on average, 0.00107185004892 of the whole space, or, in this case, 1,071.85 hard-locations.

We extended his results, calculating the distribution of the number of activated hard-locations. As each hard-location has probability $p=0.00107185004892$ of being activated, the probability of activating exactly $a$ out of $H$ hard-locations follows a binomial distribution with mean $\mu = pH$ and standard deviation $\sigma = \sqrt{Hp(p-1)}$. In this case, $\mu = 1071.85$ and $\sigma = 32.72$.

In order to validate our scan algorithm, we have run 10,000 scans from a random bitstring and counted the number of activated hard-locations. The code is available in the ``Number of activated hard-locations'' notebook \citep{sdmframework}.

In figure \ref{fig:validation-activated-hls}, we can notice that the theoretical model and the simulation matches. Hence, it seems that both the address space generation algotihm and the scan algorithm work properly. Notice that the curve is almost the same for $n=1,000$ and $n=256$. It happens because the access radius is adjust to have $p$ as close as possible to $0.001$. They are not exactly the same because their $p$ differs a little.

\begin{figure}[!htb]
  \centering
  \subfloat[$n=1,000$, $H=1,000,000$,\protect\\ $r=451$, and $p=0.00107185$ ]{\includegraphics[width=0.5\textwidth]{./images02/new-images/activated-hls-1000.png}}
  \subfloat[$n=256$, $H=1,000,000$,\protect\\ $r=103$, and $p=0.00106684$ ]{\includegraphics[width=0.5\textwidth]{./images02/new-images/activated-hls-256.png}}

  \caption{Histogram of the number of activated hard-locations in 10,000 scans from a random bitstring. The curve in red is the theoretical normal distribution with $\mu = Hp$ and $\sigma = p(p-1)H$.}
  \label{fig:validation-activated-hls}
\end{figure}

Besides the number of activated hard-locations, we have also extended Kanerva's results to calculate the distribution of distances between the center of the circle and the activated hard-locations. Let $A$ be the set of activated hard-locations, $\xi$ be the center of the circle, and $r$ be the access radius, then:

\begin{align}
P(\text{d}(a, \xi)=x | a \in A) &= \frac{P(\text{d}(a, \xi)=x)}{P(a \in A)} \\
    &= \frac{\binom{n}{x}}{\sum_{k=0}^r \binom{n}{k}} \label{eq:prob-d-inside-circle}
\end{align}

In order to check Equation \ref{eq:prob-d-inside-circle}, we have calculated the distances of the activated hard-locations to the center of 1,000 random circles. The code is available in the ``Distances of activated hard-locations'' notebook \citep{sdmframework}.

In figure \ref{fig:validation-distance-activated-hls}, we can notice that the theoretical model and the simulation matches.

\begin{figure}[!htb]
\centering\includegraphics[width=\textwidth]{./images02/new-images/distance-activated-hls.png}

\caption{Histogram of the distances of activated hard-locations to the center of the circles. The curve in red is the theoretical distribution of Equation \ref{eq:prob-d-inside-circle}
\label{fig:validation-distance-activated-hls}}
\end{figure}

\section{Intersection of two circles}

Kanerva has calculated the intersection of two circles according to the distance between their centers. The intersection is important to understand how SDM works, because it affects directly the critical distance. When $\eta_d$ is inside the critical distance, then it will converge to $\eta$. In fact, it converges because they share a sufficient amount of hard-locations, i.e., the intersection of the circle around $\eta_d$ and $\eta$ is enough to converge. For further information about the relation between the critical distance and the intersection, see \citet{brogliato2014sparse}.

We have calculated the intersection between a random bitstring (bs1) and another bitstrings (bs2) exactly $d$ bits away. The former (bs1) is just a random bitstring. The latter (bs2) was generated randomly flipping $d$ bits of bs1. The code is available in the ``Kanerva's Figure 1.2'' notebook \citep{sdmframework}.

In Figure \ref{fig:validation-intersection}, we can notice that we have obtained the same results as Kanerva. It seems that the random flipping bits algorithm and the scan algorithm work properly.

\begin{figure}[!htb]
  \centering
  \subfloat[{\citet[Figure 1.2, p.25]{Kanerva1988}} ]{\includegraphics[width=0.45\textwidth]{./images02/new-images/kanerva-table-12.jpg}}
  \subfloat[Generated by SDM-Frameworkm with $n=1,000$ ]{\includegraphics[width=0.55\textwidth]{./images02/new-images/intersection-of-circles.png}}

  \caption{Number of hard-locations in the intersection of circles around two bitstrings $x$ bits away.}
  \label{fig:validation-intersection}
\end{figure}

\section{Storage and retrieval of sequences}

\citet[Ch.8]{Kanerva1988} presented an approach to store and retrieve sequences using $k$ different SDMs, namely $\text{sdm}_1$, $\text{sdm}_2$, $dots$, $\text{sdm}_k$.

Let $a_0, a_1, a_2, \dots, a_n$ be a sequence to be stored in a $k$-fold memory. So, all pointers of the form $a_i \rightarrow a_{i+k}$ will be written to $\text{sdm}_k$ memory, i.e., in $\text{sdm}_1$, the following pointers will be written: $a_0 \rightarrow a_1$, $a_1 \rightarrow a_2$, $\dots$, $a_{n-1} \rightarrow a_n$; while in $\text{sdm}_2$, the following pointers will be written: $a_0 \rightarrow a_2$, $a_2 \rightarrow a_3$, $\dots$, $a_{n-2} \rightarrow a_n$; and so forth.

We have tested exactly the same example presented in \citet{Kanerva1988}, p.85. We wrote two sequences to a $3$-fold memory: $<A, B, C, D>$ and $<E, B, C, F>$. Then, after reading the sequences $<A, B, C>$ and $<E, B, C>$, we have obtained $D$ and $F$, respectively.

Each reading operation was performed summing the counters of all activated hard-locations from all three memories. For instance, to read the sequence $<A, B, C>$, we have activated the hard-locations around $C$ in $\text{sdm}_1$, we have also activated the hard-locations around $D$ in $\text{sdm}_2$, and, finally, we have also activated the hard-locations around $A$ in $\text{sdm}_3$. After summing the counters of all those hard-locations, we evaluate the resulting bitstring just as in the original read operation.

The code is available in the ``Sequences (Kanerva Ch 8)'' notebook \citep{sdmframework}.

The logic behind how it works is that, when reading the sequence $<A, B, C>$, we have $A$ pointing to $D$, while both $B$ and $C$ point to $D$ and $F$. Thus, $D$ appears more often than $F$ and ended up being the result.

Hence, as we have replicated the theoretical results from Kanerva, we have one more evidence that our framework works properly.

\subsection{$k$-fold memory using only one SDM}

We have extended Kanerva's ideas to be able to store and retrieve sequences in $k$-fold memories using only one SDM (instead of $k$ SDMs).

Our idea was to create $k$ random bitstrings, one for each fold. We have performed writing and reading exactly as Kanerva's original idea, but, instead of writing to $\text{sdm}_k$, we have written $a_{i+k}$ into the address $a_i \oplus \text{tag}_k$, and, instead of reading from $\text{sdm}_k$, we have read from address $a_i \oplus \text{tag}_k$, where $\oplus$ is the exclusive or (XOR) operator.

It worked as if we had splitted SDM into $k$ regions with low intersection between two of them. So, as the interference is minimal, they work like independent SDMs. The major disadvantage of this approach is that memory capacity may be reached faster.

Splitting the memory into regions may be an interesting strategy to other sorts of problems, mostly the ones which would need many SDMs and, consequently, would use a lot of RAM.

\section{Convergence of $\eta_x$ to $\eta$}

One particular analysis of Kanerva's interest is that of the distance read at a point $\eta_x$. Suppose an SDM is trying to read an item written at $\eta$, but the cues received so far lead to a point of distance $x$ from $\eta$.  As one reads at $\eta_x$, a new bitstring $\beta$ is obtained, leading to Kanerva's question: what is the new distance from $\eta$ to $\beta$? Is it smaller or larger than $x$? That, of course, depends on the ratio between $x$ and the number of dimensions of the memory.

\citet[p.70]{Kanerva1988} originally predicted a \textasciitilde 500-bit distance after a point (Figure \ref{fig:kanerva-figure-7.3}). The original prediction considered that the read distance would decline when inside the critical distance and increase afterwards, converging to a \textasciitilde 500-bit distance.  At this point, each read would lead to a different, orthogonal, \textasciitilde 500-bit distance bitstring. He analyzed specifically an SDM with 1,000 bits and 10,000 random bitstrings written into it.

\begin{figure}[h]
\centering\includegraphics[width=0.8\textwidth]{images02/kanerva-table-7-2-original.png}
\caption{Kanerva's original Figure 7.3 (p. 70) predicting a \textasciitilde 500-bit distance after a point.
\label{fig:kanerva-figure-7.3}}
\end{figure}

As we ran the simulations, this one in particular struck our attention: The new distances obtained after a read operation were not perfectly predicted by the theoretical model.
%and we propose that this is due to interaction effects between different attractors.

We have strictly followed Kanerva's configuration and, even so, we have found out some deviations from Kanerva's original theoretical analysis and the results obtained by simulation.

In details, we have created a SDM with $n=1,000$, $H=1,000,000$, and $r=451$. Then, we have generated 10,000 random bitstrings and written them into the memory. Then, we have generated a reference bitstring (bs\_ref) and written it into the memory. Then, we have executed the following steps with $x$ from 0 to 1,000: (i) copy bs\_ref into a new bitstring; (ii) randomly flipped $x$ bits of the copy; (iii) read from the memory in the copy address; and (iv) stored the distance between the returned bitstring and bs\_ref. Finally, we have plotted Figure \ref{fig:sdm-10000w-table-7-2}.

\begin{figure}[h]
\centering
\subfloat[1 sample for each distance $x$ \label{fig:sdm-10000w-table-7-2-1sample} ]{\includegraphics[width=0.5\textwidth]{./images02/sdm-10000w-table-7-2.png}}
\subfloat[6 samples for each distance $x$ \label{fig:sdm-10000w-table-7-2-6samples} ]{\includegraphics[width=0.5\textwidth]{./images02/sdm-10000w-table-7-2-6-samples.png}}

\caption{Results generated by the framework diverging from Kanerva's original Table 7.2. Here we had a 1,000 bit, 1,000,000 hard-location SDM with exactly 10,000 random bitstrings written into it, which was also Kanerva's configuration.
\label{fig:sdm-10000w-table-7-2}}
\end{figure}

Figure \ref{fig:sdm-10000w-table-7-2-1sample} has a lot of noise because we have read only once for each distance $x$ and Kanerva has predicted the average distance. So, we have changed the steps to run $k$ reads and store the average new distance. We run with $k=6$, and the results can be seen in Figure \ref{fig:sdm-10000w-table-7-2-6samples}, which has a way lower noise and still holds the divergence.

Our results show that the theoretical prediction is not accurate.  There are interaction effects from one or more of the attractors created by the 10,000 writes, and these attractors seem to raise the distance beyond \textasciitilde 500 bits (Figure \ref{fig:sdm-10000w-table-7-2}).

Obviously, these small deviations from Kanerva's original theoretical predictions deserve a qualification.  Kanerva was working in the 1980s and the 1990s, and had no access to the immense computational power that we do today. It is no surprise that some small interaction effects should exist as machines allow us to explore the ideas of his monumental work.

But, when we reduced the number of random bitstrings written in the SDM from 10,000 to only 100, the results reflected very well the Kanerva's theoretical expectation (Figure \ref{fig:sdm-100w-table-7-2-10samples}). This result strengthens our hypothesis that the disparities in the computational results are due to the interaction effect of high numbers of different attractors. In Figure \ref{fig:sdm-table-7-2-steps} we can notice that, the more random bitstrings are written, the stronger the attractors.

\begin{figure}[h]
\centering
\subfloat[100 writes \label{fig:sdm-100w-table-7-2-10samples} ]{\includegraphics[width=0.47\textwidth]{./images02/sdm-100w-table-7-2.png}}
\subfloat[Steps of 1,000 writes \label{fig:sdm-table-7-2-steps} ]{\includegraphics[width=0.53\textwidth]{./images02/sdm-table-7-2-steps.png}}
%\centering\includegraphics[width=\textwidth]{images02/sdm-100w-table-7-2.png}

\caption{Results generated by the framework similar from Kanerva's original Table 7.2. Here we have a 1,000 bit, 1,000,000 hard-location SDM with (a) exactly 100 random bitstrings written into it and (b) steps of 1,000 random bitstrings written into it.
\label{fig:sdm-100w-table-7-2}}
\end{figure}

To obtain the results from Figures \ref{fig:sdm-10000w-table-7-2} and \ref{fig:sdm-100w-table-7-2}, we had to write 10,000 random bitstrings to an SDM, and then randomly choose one of those bitstrings to be our origin. Finally, we randomly flipped some bits from the origin bitstring and executed a reading operation in the SDM. Thereby, in order to show the interaction effects more clearly, we changed the single read for an 15-iterative read. As we can see in Figure \ref{sdm-10000w-table-7-2-15iter}, after a distance of 500 bits, all bitstrings converged to 500-bit distance bitstrings, just as described by Kanerva.

Hence, our understanding is that the attractors are just preventing the bitstrings to converge directly to 500-bit distance bitstrings, requiring more reading steps to do so. They are in other orthogonal bitstrings' critical distance, but sufficiently far not to converge in a single read.

\begin{figure}[h]
\centering\includegraphics[width=\textwidth]{images02/sdm-10000w-table-7-2-15iter.png}
\caption{This graph shows the interaction effects more clearly.  As we change the single read to a 6-iterative read, the effect has vanished and all bitstrings above $x=500$ have converged to 500-bit distance bitstrings. Here we have the exact same configuration of Figure \ref{sdm-10000w-table-7-2}, except for the iterative read.
\label{sdm-10000w-table-7-2-15iter}}
\end{figure}

%To obtain the results from Figures \ref{fig:sdm-10000w-table-7-2} and \ref{fig:sdm-100w-table-7-2}, we had to write 10,000 random bitstrings to an SDM, and then randomly choose one of those bitstrings to be our origin. Finally, we randomly flipped some bits from the origin bitstring and executed a reading operation in the SDM. Thereby, in order to show the interaction effects more clearly, we wrote a handmade bitstring to the SDM which had all bits inverted in relation to the origin bitstring --- their hamming distance was equal to 1,000. Our handmade bitstring was acting as an opposite attractor, and one can see the accelerating effects towards convergence to both attractors: the origin and the handmade bitstrings (Fig. \ref{sdm-10000w-notX-table-7-2}). Here we had the exact same configuration of Figure \ref{sdm-10000w-table-7-2}, with the addition of the single opposite attractor.

%\begin{figure}[h]
%\centering\includegraphics[width=0.8\textwidth]{images02/sdm-10000w-notX-table-7-2.png}
%\caption{This graph shows the interaction effects more clearly.  As we include an opposite bitstring, one can see the accelerating effects towards convergence to both attractors: the origin and the opposite. Here we have the exact same configuration of Figure \ref{sdm-10000w-table-7-2}, with the addition of the single opposite attractor.
%\label{sdm-10000w-notX-table-7-2}}
%\end{figure}

\chapter{Results (ii): Loss of neurons}

In SDM, the data is written distributed among millions of hard-locations, which theoretically gives SDM robustness against loss of neurons. In other words, SDM should keep convering correctly even when some neurons are dead. The question is: how robust it really is? How many neurons may die before it starts to forget things. These questions have never been addresses before.

@@

\chapter{Results (iv): Performance}

Performance is an important part of our framework. Running an SDM memory consumes a lot of resource from both CPU and RAM.

For instance, each scan in 1,000,000 hard-locations with 1,000 bits executes $10^9$ bit compares through $10^9/64 = 15,625,000$ XORs and calls to the built-in popcount. So, 10,000 writes execute $10^{13}$ bit compares, while a 6-iterative reading executes $6 \cdot 10^{12}$ bit compares. The heatmap of Figure \ref{fig:cir-dist-10k-writes-kanerva} executed $3.05 \cdot 10^{15}$ bit compares. For comparison, the number of seconds since Jesus's birth is 63,639,648,000 = $6.36 \cdot 10^{10}$. The number of people who have ever lived on earth is estimated to be $1,08 \cdot 10^{11}$. There are approximtely $1.8 \cdot 10^{9}$ websites in the internet.

Amazon EC2 p3.2xlarge has generated the heatmap of Figure \ref{fig:cir-dist-10k-writes-kanerva} in 15 minutes and 3 seconds. It has compared $3.37 \cdot 10^{12}$ bits per second through $52.6 \cdot 10^{9} = 52.6 \text{ billion}$ XORs and popcounts per second.

We have executed the same performance test in each device. The test has 3 parts. The first part consists of: 1,000 linear scans, 5,000 thread scans, and 5,000 OpenCL scans. The second part consists of 5,000 writes, 5,000 single reads, and 1,000 6-iterative readings, with both thread and opencl scanners. The code is available in the ``Performance test'' notebook \citep{sdmframework}.

Our first device is a personal MacBook Pro Retina 13-inch Late 2013 with a 2.6GHz Intel core i5 processor, 6GB DDR3 RAM, and Intel Iris GPU.

Our second device is an iMac Retina 5K 27-inch 2017 with a 3.8GHz Intel core i5 processor, 8GB DDR4 RAM, and a Radeon Pro 580 8G CPU. The complete results are presented in Figures \ref{fig:perf-imac-read-256}, \ref{fig:perf-imac-write-256}, \ref{fig:perf-imac-read-1000}, \ref{fig:perf-imac-write-1000}, \ref{fig:perf-imac-scan-256}, \ref{fig:perf-imac-scan-1000}, and \ref{fig:perf-imac-scan-10k} and Tables \ref{tab:perf-imac-256}, \ref{tab:perf-imac-1000}, and \ref{tab:perf-imac-10k}.

Beyond that, we are also running in state-of-the-art devices: (i) an Amazon EC2 p2.xlarge with Intel Xeon E5-2686v4 processor, 61GB DDR3 RAM, and NVIDIA K80 GPU, and (ii) an Amazon EC2 p3.2xlarge with Intel Xeon E5-2686v4 processor, 488GB DDR3 RAM, and NVIDIA Tesla V100 GPU.

It is interesting that which scanner is faster depends also on the SDM settings. In the iMac 2017, the faster scanner for a 1,000-bits SDM was the OpenCL, but for a 256-bit SDM was the threads. What happened here is that the OpenCL kernel chosen was a generic one which performs the scan for any SDM. It is always possible to optimize the OpenCL kernel to a specific setting, and it would be faster than the threads. By default, the framework chooses a generic kernel which we believe would be reasonable for the most common setups.
% TODO Run the optimized kernel and show the results.

Most of the time, the bottleneck of both the read and the write operations is the scanner. But we have optimized the OpenCL kernel so much for the iMac 2017 that scanner not the bottleneck anymore. In the writing operation, it took the same time to scan and to update the counters. It is impressive because it update, on average, 1,000 counters of 32-bit integers each, and it was as fast as (i) sending the command to the GPU, (ii) performing 1 billion bit compares, and (iii) downloading the response from the GPU.

Our conclusion is that, if one is really concerned about performance, one should fine tune the OpenCL kernel for one's GPU. It would always be faster than running in the CPU.

\begin{figure}[!htb]
\centering\includegraphics[width=\textwidth]{images02/performance/imac-read-256.png}
\caption{Time to run a single read in a SDM with $n=256$, $H=1,000,000$ and $r=103$.
\label{fig:perf-imac-read-256}}
\end{figure}

\begin{figure}[!htb]
\centering\includegraphics[width=\textwidth]{images02/performance/imac-write-256.png}
\caption{Time to run one write in a SDM with $n=256$, $H=1,000,000$ and $r=103$.
\label{fig:perf-imac-write-256}}
\end{figure}

\begin{figure}[!htb]
\centering\includegraphics[width=\textwidth]{images02/performance/imac-write-1000.png}
\caption{Time to run one write in a SDM with $n=1,000$, $H=1,000,000$ and $r=451$.
\label{fig:perf-imac-write-1000}}
\end{figure}

\begin{figure}[!htb]
\centering\includegraphics[width=\textwidth]{images02/performance/imac-read-1000.png}
\caption{Time to run a single read in a SDM with $n=1,000$, $H=1,000,000$ and $r=451$.
\label{fig:perf-imac-read-1000}}
\end{figure}

\begin{figure}[!htb]
\centering\includegraphics[width=\textwidth]{images02/performance/imac-scans-256.png}
\caption{Time to run a single scan in a SDM with $n=256$, $H=1,000,000$ and $r=103$.
\label{fig:perf-imac-scan-256}}
\end{figure}

\begin{figure}[!htb]
\centering\includegraphics[width=\textwidth]{images02/performance/imac-scans-1000.png}
\caption{Time to run a single scan in a SDM with $n=1,000$, $H=1,000,000$ and $r=451$.
\label{fig:perf-imac-scan-1000}}
\end{figure}

\begin{figure}[!htb]
\centering\includegraphics[width=\textwidth]{images02/performance/imac-scans-10k.png}
\caption{Time to run a single scan in a SDM with $n=10,000$, $H=1,000,000$ and $r=4845$.
\label{fig:perf-imac-scan-10k}}
\end{figure}

>>>>

\chapter{Sparse Distributed Memory}
\input{./Chapters/sdm2.tex}

\chapter{Framework Architecture}
\input{./Chapters/sdm-architecture.tex}

\chapter{Results (i): Framework Validation}
\input{./Chapters/sdm-validation.tex}

\chapter{Results (ii): Loss of neurons}
In SDM, the data is written distributed among millions of hard-locations, which theoretically gives SDM robustness against loss of neurons. In other words, SDM should keep convering correctly even when some neurons are dead. The question is: how robust it really is? How many neurons may die before it starts to forget things. These questions have never been addresses before.

@@

\chapter{Results (iv): Performance}
\input{./Chapters/sdm-performance.tex}

% Application
\input{./Chapters/sdm-applications.tex}
